Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-07 text/manuscript-full-KS.qmd
full version copy

```{=tex}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Data and Methods

## Data and Variables

**RQ1**. Our study utilizes two main data sources based on each research question. The first research question is focused on understanding the socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tract is the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Census tracts are statistical subdivisions of a county designed to capture relatively homogeneous demographic communities that are generally considered to be more accurate for population analyses than zip codes, which are spatial boundaries drawn to demarcate route efficiency for postal services \[[CITE](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html)\] .

For the first research question, we use data from the 1980 and 2000 Decennial Census as well as 5-year estimates from the 2020 American Community Survey (ACS). For the 1980 Decennial Census, we specifically draw on Summary Tape File 1 (STF1) for variables about race and Hispanic origin collected via “short form” questions answered by all households and Summary Tape File 3 (STF3) for variables about socioeconomic characteristics (income, education, and poverty) collected via the "long form" questionnaire completed by a sample of households. The 1980 Decennial Census data collection period closely matches the data collection period of SAT score-sending behavior that @RN4982 utilized to create Geomarkets and the Market Segment Model. To examine how Geomarket characteristics change over time, we similarly use the Summary File 1A (SF1A) for census-tract measures of race and Hispanic origin and the Summary File 3A (SF3A) for census-tract measures of income, education, and poverty from the 2000 Decennial Census.[^1] Beginning in 2010, the Decennial Census no longer collected a "long form" questionnaire. These questions were replaced by ACS. Therefore, we also utilized the 2020 5-year ACS for both race and socioeconomic characteristics at the census tract level, which includes data collected from 2016-2020.[^2] This data collection period mirrors the period for our primary data collection about student list purchases used to analyze RQ2, as described below.

[^1]: Data for the 1980 and 2000 Decennial Census were retrieved from the IPUMS National Historical Geographic Information System, created by @RN5028.

[^2]: Data for the 2020 5-year ACS were retreived from the [tidycensus](https://walker-data.com/tidycensus/) R package, created by @tidycensus_citation

Across these data sources, we created measures of race and ethnicity, specifically the number and proportion of the census-tract population in the following categories: White, non-Hispanic; Black, non-Hispanic; Asian, non-Hispanic; Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two or more races, non-Hispanic. Measures of Asian, American Indian and Alaskan Native, two or more races, and Native Hawaiian and Pacific Islander (all non-Hispanic) were not available for the 1980 Decennial Census. We also created socioeconomic measures that approximate the predictors of market segment identified by @RN4982. These variables are median household income, percent of households below the poverty line, and people age 25+ with a BA. All measures were created at the census tract-level and the Geomarket-level by aggregating tract-level data to Geomarkets.

We assigned each census tract to College Board Geomarkets by implementing a partial spatial merge. Spatial manipulations utilized functions from the *sf* R package. Data to create the shapefiles for the College Board Geomarkets were obtained from a 2012 R-bloggers post.[^3] Essentially, the data file had one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the `aggregate()` function to create shapefiles for each Geomarket.

[^3]: The author of the post reported that he found the zip codes associated with Geomarkets from a Google search.

Next, we utilized the `st_intersection()` function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. Census tract shapefiles that intersected with multiple Geomarkets (i.e., a census tract's spatial boundaries were spread across two or more Geomarkets) broken into smaller shapefiles wholly contained within a single Geomarket. The variables in these shapefiles were then replaced with the original variable value multiplied by the proportion of land area of the census tract that was contained within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 1,000 people who identify as Hispanic non-white. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*1000 = 600$ Hispanic non-white people and the observation assigned to Geomarket B reports $.40*1000 = 400$ Hispanic non-white people.

**RQ2**. Our study's second research question asks how does the racial and socioeconomic composition of included versus included prospects vary when student list purchases filter on Geomarkets. We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from the search criteria.

The data collection sample for this project was all public universities in California, Illinois, Minnesota, and Texas. Utilizing public records requests to obtain public records is a painstaking process. Initially, the majority of universities did not respond or denied our request. Subsequently, we obtained pro bono representation from four law firms, which substantially increased the success of data collection. We also collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for Texas.

Even with firm representation, data collection remained difficult. Some universities provided records that were not usable for quantitative analyses (e.g., summary statistics across multiple orders; or data did not contain important fields). Some universities did not provide records based on legitimate legal grounds (e.g., data not in university possession; not required to create records that do not currently exist). We learned that many universities outsourced student list purchases to third-party consulting firms. Unfortunately, we were rarely able to obtain usable data from these universities. A small number of universities denied requests based on potentially questionable legal rationale, but we lacked the resources to litigate.

The primary data source for our second research question is therefore student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 \[KS CHECK THIS\] public universities located in California, Illinois, Minnesota, and Arizona \[KS Does this still apply?\]. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase.[^4] The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters. Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

[^4]: ADD CAVEAT THAT PURCHASES EXCLUDE PROSPECT PROFILES THAT WERE PREVIOUSLY PURCHASED

In particular, we answer the second research question by analyzing particular student list purchases that did *not* utilize the Geomarket search filter, but did purchase prospects from all Geomarkets in a particular metropolitan area. For each prospect we know their home zip code and, therefore, which Geomarket they belong to. We therefore simulate which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets.

The variables of interest for student list data are the characteristics of prospects who would have been included versus excluded by our simulations by Geomarket filter. Prospect characteristics are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity nonresponse) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want (College Board, 2016). From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic (College Board, 2016).[^5] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire \[KS double check this\].

[^5]: Any student who selects a Hispanic ethnicity category is defined as Hispanic/Latino, regardless of the race categories they select. Additionally, non-Hispanic students who check “American Indian or Alaska Native” and another race group are defined as “two or more races, non-Hispanic.”

## Methods

**Research design**. We utilize a multiple, quantitative case study design in which metropolitan areas are cases. Case studies are used to gain a deeper understanding of the phenomenon of interest by comparing and contrasting across different settings \[CITE YIN 2001\]. Our study's research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarket landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We therefore focus on minimizing the number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

Case selection was therefore informed by purposeful sampling to identify “information-rich” cases (Patton, 2002; Yin, 2009) that can provide in-depth knowledge about variation across Geomarkets and overtime. We considered metropolitan areas as information rich cases across three broad dimensions: 1) urban model structures, 2) geographical variation, 3) variation across racial/ethnic and socioeconomic characteristics, and 4) availability of student list orders and de-identified list data. Based on our read of @RN4982, we identified classic urban model structures that reflected "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12), such as Boston, Chicago, Detroit, Baltimore, Denver, Houston, Miami, Philadelphia**,** Phoenix, and Pittsburgh. To compare and contrast with classic urban model structures, we also identified alternative urban models such as bimodal or metroplex models that contain two centralized cores surrounded by the more classing suburban ring (e.g., Dallas-Fort Worth and San Francisco) and multi-nuclei models where urban growth develops around multiple, sometimes uncentralized and independent, metropolitan areas (e.g., Los Angeles and New York City). The third consideration addresses geographical variations in demographic populations by intentionally selecting metropolitan cases across different U.S. regions (West, Midwest, Northeast, South) to ensure resident and prospect pools mirror broader populations. Cumulatively related to all previous selection dimensions, we are also interested in variation of racial and socioeconomic populations between Geomarkets within a metropolitan areas (RQ1) and how such variation shapes which prospective students are targeted by student list purchases (RQ2). We therefore considered whether or not metropolitan areas provided sufficient and varied racial and socioeconomic diversity for our research purposes.

Lastly, choice of metropolitan cases was informed by student list data availability. Although Census data are available for all metropolitan areas nationally (RQ1), we do not have good candidate student list data for all metropolitan areas (RQ2). In order to simulate which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets, student list data must contain orders that resulted in a large pool of prospects within the metropolitan areas considered above that would include all Geomarkets within metropolitan areas. This lead us to select on orders that filtered for an entire state (e.g., State filter is set to Pennsylvania), an entire metropolitan area without a Geomarket filter (CBSA filter is set to the Philadelphia metropolitan area) , or an entire metropolitan area with a Geomarket filter inclusive of all Geomarkets (CBSA filter is set to the Philadelphia metropolitan area and Geomarket filter is set to all five Geomarkets in the Philadelphia area). In order to isolate the impact of Geomarkets on the racial/ethnic and socioeconomic characteristics of prospect pools, a suitable metropolitan case candidate would also be contingent on student list orders that did not include too many filtering criteria beyond those detailed previously. Because nearly all orders we collected (Author, BLINDED) filter on some academic criteria, we only considered orders that included assessment threshold filters (e.g., PSAT, SAT) across a wide score range to explore whether and to what extent Geomarket contributions to prospect pools are consistent across all score ranges.

Our case selection process resulted in three metropolitan cases: Chicago, Dallas-Fort Worth, and Los Angeles.

**Analyses**. Analyses are simultaneously descriptive and spatial. We answer RQ1 by producing Geomarket-level tables and graphs that show how the Geomarkets in a selected metropolitan area vary on racial and socioeconomic characteristics and how they vary over time. We also produce interactive maps at the census tract-level to show more granular variation within and between Geomarkets.

We answer RQ2 by analyzing student list purchases that encompassed all Geomarkets in a selected metropolitan areas. Descriptive tables and graphs describe the racial and socioeconomic characteristics of prospects that would have been included/excluded had the purchase filtered on particular Geomarkets. We also create interactive maps of prospect pools by Geomarket contributions to provide a more granular visualization of analyses.

```{=html}

<!-- [DO WE NEED THIS LEVEL OF DETAIL???]

For example, for purchased profiles who lived in the Chester County Geomarket (PA02) [OR PHILADELPHIA PA05] and identified as Black, where in the county did these students live and which high schools did they attend?

 For example, for purchased profiles who lived in the Chester County Geomarket (PA02) [OR PHILADELPHIA PA05] and identified as Black, where in the county did these students live and which high schools did they attend?

NEXT STEPS: CREATE TABLE FOR PHILLY STUDENT LIST PURCHASES THAT FOCUSES ON FIRST GEN STATUS.

Our conceptual framework suggests that racial and socioeconomic inequality is Geomarket regions is associated with the number of geomarkets in the metropolitan area. and is associated with the patterns of racial segregation in the metropolitan area. comment more on this.

Below, we use simulations and actual student list purchases that filtered on geomarkets. In purchases that filtered on test-score and/or GPA thresholds, we can simluate who would be included and excluded had certain Geomarkets were been selected. In purchases that filter on geomarkets, we can get an initial sense of who is included in geomarkets targeted by regional state colleges versus research universities.

-->
```

>>>>

```{=tex}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Introduction

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

[^1]: On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.

```{=html}
<!--Geomarkets carve states and metroplitan areas into smaller geographic units and have names like,  On the same day, the university purchased two additional orders that were identical, except they filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$. 
On the same day, the university also purchased orders $\#448427$ ($57,537$ names) and $\#448440$ ($78,000$ names), which filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets as order $\#448922$.

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```
Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary for their progeny to claim spots at selective colleges [@RN1228; @RN1241; @RN4991] [ADD CITES]. @RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [@RN861] \[cite others\]. Scholarship on enrollment management offers an agentic depiction of organizational behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

We argue that these perspectives ignore an important mechansim of social reproduction: third-party vendors create products that sort students on behalf of college. These products often amplify historic inequality in educational opportunity by using past snapshots of student demand to make recommendations about where colleges should prioritize recruiting efforts. In the contemporary era of private equity funded platform capitalism, third-party recruiting products have become increasingly powerful, granular, and ubiquitous. However, we argue that a watershed moment in their develeopment was the creation of Geomarkets by the College Board and UPenn professor, Robert Zemsky.

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board underwrote the project -- providing data and funding -- and published the book as part of their nascent efforts "to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. The broader project was conceptualized as an effort to to "capture and quantify" (p. 11) the knowledge of admissions officers, based on the idea that "a good recruiter knows where to look for prospective applicants, as seen in the students' willingness or reluctance to travel" (p. 11). @RN4982 developed the Market Segment Model by analyzing the SAT score-sending behavior of 1980 high school seniors. The thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets, a territory of an admissions recruiter. Figure X, panel A shows Geomarkets in the Chicago-land area. As a planning tool for colleges, the Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

Geomarkets have been incorporated into the supply-side structure of college access in three ways. Anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. Figure X panel B shows that Chicago-land Geomarkets define recruiting territories for University of Chicago admissions recruiters (Appendix Figures BLAH BLAH). Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), which was in 1984 and remains active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. According to @RN5021, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis product, which was based on EPS.

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz]. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets.

Scholarship on enrollment management has focused on college behavior [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. Scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with sorting [@RN4873], particularly sorting into college.

This manuscript analyzes College Board Geomarkets as a case study of quantification. In particular, we draw from the discussions of correlation and homophily from @RN4975. @RN4982 identified the correlates of 1980 SAT score-sending behavior and concluded that student demand for higher education is primarily a function of social class. The Market Segment Model argues that homophily -- actors that share characteristics form connections -- is the organizing principle of competition and college choice, stating that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Scholarship on quantification demonstrates that making recommendations based on past correlations amplifies the effects of historic structural inequality [@RN4794]. The snapshot of student demand in 1980 -- itself a consequence historic structural inequality -- was programmed into recruiting products that colleges utilize to identify and target prospective students. The result is a supply-side that amplifies structural inequalities observed on the demand-side.

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? This research question is motivated by @RN4982, which recommends that selective colleges target affluent Geomarkets. Additionally, prior research on enrollment management behavior finds that selective colleges focus recruiting visits on high schools in affluent, predominantly white communities [@RN3519; @RN4758]. We analyze this research question using data from actual student list purchases, which were collected using public records requests. We analyze student list purchases that include all Geomarkets in a metropolitan area in only to assess which prospects would be included/excluded by had the student list purchase filtered on particular Geomarkets.

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification, particularly correlation and homophily. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. [We find that\ldots ?PARAGRAPH ON RESULTS?] Finally, we discuss implications for scholarship and for policy. The sociology of education should move beyond the fixation on schools and policies as the primary sorting machines. @RN527 [p. 148] remind us that the organizational field includes the totality of relevant actors." Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product \[CITE\]. By investigating high-leverage actors and products in the organizational field, sociology can develop important insights that shape future federal and state regulatory policy. \[REVISE THIS PARAGRAPH\]

```{=html}
\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```
# Enrollment Management and Quantification

**Enrollment Management**

<!-- la la -->

Enrollment management is simultaneously a profession, an administrative structure, and an industry. As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  -->

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Scholarship from the economics of education often investigate financial aid leveraging, which seeks to convert admits to enrolled students [e.g., @RN3762; @RN3133]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. In turn, recruiting visits to high schools -- and subsequent phone calls -- are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly white public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

By focusing on the behavior of colleges, scholarship on EM implicitly states that recruiting is something done by individual colleges. The broader EM industry includes relevant stakeholders in the organizational field [@RN527], including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges. We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways enrollment management contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  --> <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory, enrollment management processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. Unfortunately, few empirical studies make EM vendors or products the object of empirical analysis. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

[^2]: In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The Student Search Service -- created in 1972 and copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. Second, the Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification. Student Search Service includes or excludes prospects from a student list purchase based on value of selected variables while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on equality of opportunity.

```{=html}
<!--

The EM industry puts great stock in market research, in part because the EM profession emerged from marketing and market research [@RN4986; @RN2771]. @RN4986 [p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].[^1] In recruiting, market research informs decisions about which prospective students to target with recruiting interventions. How to acquire market research is another make-or-buy decision. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In 1983, College Board published *Applying Market Research in College Admissions* [@RN4986], which became the basis for their Admitted Student Questionnaire (ASQ) product, and published *The Structure of College Choice* [@RN4982], which became the basis for the Enrollment Planning Service (EPS) product/division. Today, EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Market research often distills large amounts of data into simpler indicators. Therefore, the quantification literature provides a useful lens to understand market research.

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->
```
**Quantification**

@RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402). Reacivity, discipline, and authority are three interrelated themes that describe the effects of quantification. *Reactivity* is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification *disciplines* actors to react in particular ways. Quantification changes decision-making power *authority*, often weakening the discretion of local decision-makers. Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions personnel to make independent decisions class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->
